---
title: "Replication of Karlan & List (2007)"
subtitle: "Does Price Matter in Charitable Giving? Evidence from a Large-Scale Natural Field Experiment"
author: "Riley Li"
date: today
format:
  html:
    toc: true
    code-fold: true
    fig-width: 7
    fig-height: 4
    fig-align: center
execute:
  warning: false
  message: false
---

```{r}
#| echo: false
library(haven)
library(tidyverse)
library(sandwich)
library(lmtest)
library(knitr)
library(kableExtra)

df <- read_dta("Data/AERtables1-5.dta")
```


## Introduction

Does making a donation cheaper — by having a third party match it — actually change whether people give, or how much they give? This is fundamentally a question about **price elasticity of charitable giving**.

Karlan & List (2007) answer this question using a large-scale natural field experiment conducted with a U.S. liberal political advocacy organization. They mailed approximately 50,000 prior donors one of several solicitation letters. The **treatment group** was offered a matching grant (a lead donor would match every dollar raised at either a 1:1, 2:1, or 3:1 ratio, up to some threshold), effectively reducing the "price" of giving. The **control group** received a standard solicitation letter with no matching offer.

By randomly assigning donors to treatment and control, the experiment identifies a causal effect of the match offer. The key findings are:

- Simply announcing a matching grant **significantly increases** both the likelihood and the total amount of charitable giving.
- Larger match ratios (2:1, 3:1) do **not** improve giving beyond a 1:1 match — suggesting that the mere existence of a match matters more than its size.
- Political partisanship moderates the effect: matching grants are more effective in "red" (Republican-leaning) states.

The experiment provides some of the cleanest causal evidence available on how price incentives shape philanthropic behavior.


## Table 1: Checking for Selection Bias

A critical assumption for any randomized experiment is that the treatment and control groups are **balanced on pre-treatment characteristics** — that is, randomization worked. Table 1 below replicates the balance check from the paper.

For each baseline variable, I report the mean in the control group, the mean in the treatment group, and a p-value from a two-sample t-test. A p-value above 0.05 means we cannot reject the null of equal means — evidence of successful randomization.

```{r}
#| label: table1

# Variables used in Table 1 of the paper
t1_vars <- c("red0", "redcty", "nonlit", "cases", "MRM2", "HPA",
             "freq", "years", "dormant", "female", "couple")

t1_labels <- c(
  red0    = "Respondent in red state (2004)",
  redcty  = "Respondent in red county (2004)",
  nonlit  = "No. months non-literate",
  cases   = "No. months given in past 2 years",
  MRM2    = "Months since last gift",
  HPA     = "Highest previous contribution ($)",
  freq    = "Frequency of giving (gifts/year)",
  years   = "Years on mailing list",
  dormant = "Dormant donor (2+ years since last gift)",
  female  = "Female",
  couple  = "Couple"
)

t1_results <- map_dfr(t1_vars, function(v) {
  ctrl <- df[[v]][df$control == 1 & !is.na(df[[v]])]
  trt  <- df[[v]][df$treatment == 1 & !is.na(df[[v]])]
  tt   <- t.test(ctrl, trt)
  tibble(
    Variable   = t1_labels[v],
    Control    = round(mean(ctrl, na.rm = TRUE), 3),
    Treatment  = round(mean(trt,  na.rm = TRUE), 3),
    Difference = round(mean(trt, na.rm = TRUE) - mean(ctrl, na.rm = TRUE), 3),
    `p-value`  = round(tt$p.value, 3)
  )
})

kable(
  t1_results,
  caption = "Table 1: Pre-Treatment Balance Check (Treatment vs. Control)",
  align   = c("l", "r", "r", "r", "r")
) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) |>
  add_header_above(c(" " = 1, "Group Means" = 2, " " = 2))
```

No individual variable shows a statistically significant difference at conventional levels (p < 0.05), consistent with successful randomization. This confirms there is **no meaningful selection bias** between treatment and control groups — any differences we observe in outcomes can be attributed to the matching grant offer itself.


## Table 2A (Panel A): Summary Statistics by Experimental Group

Panel A of Table 2 reports three outcome measures across treatment arms:

1. **Response rate**: the share of recipients who donated at all (`gave`)
2. **Amount given (unconditional)**: average donation across *all* recipients, including non-donors
3. **Amount given (conditional)**: average donation among *donors only*

```{r}
#| label: table2a

# Helper: compute mean, se, n for a variable within a subset
summarize_var <- function(data, var, condition) {
  x <- data[[var]][condition & !is.na(data[[var]])]
  tibble(
    n    = length(x),
    mean = mean(x),
    se   = sd(x) / sqrt(length(x))
  )
}

groups <- list(
  "Control"        = df$control == 1,
  "Treatment"      = df$treatment == 1,
  "Ratio 1:1"      = df$ratio == 1  & !is.na(df$ratio),
  "Ratio 2:1"      = df$ratio == 2  & !is.na(df$ratio),
  "Ratio 3:1"      = df$ratio == 3  & !is.na(df$ratio)
)

build_panel <- function(outcome_var, condition_extra = NULL) {
  map_dfr(names(groups), function(g) {
    cond <- groups[[g]]
    if (!is.null(condition_extra)) cond <- cond & condition_extra
    s <- summarize_var(df, outcome_var, cond)
    tibble(Group = g, N = s$n,
           Mean = round(s$mean, 4),
           SE   = round(s$se, 4))
  })
}

gave_panel   <- build_panel("gave")
amount_panel <- build_panel("amount")
cond_panel   <- build_panel("amount", df$gave == 1)

t2 <- gave_panel |>
  rename(`Response Rate` = Mean, `SE (RR)` = SE) |>
  left_join(amount_panel |> select(Group, `Amt Uncond.` = Mean, `SE (Unc)` = SE), by = "Group") |>
  left_join(cond_panel   |> select(Group, `Amt Cond.`  = Mean, `SE (Con)` = SE), by = "Group") |>
  select(Group, N, `Response Rate`, `SE (RR)`, `Amt Uncond.`, `SE (Unc)`, `Amt Cond.`, `SE (Con)`)

kable(
  t2,
  caption = "Table 2A (Panel A): Response Rates and Donation Amounts by Group",
  align   = c("l", "r", "r", "r", "r", "r", "r", "r")
) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) |>
  add_header_above(c(" " = 2, "Response Rate" = 2, "Amount (Uncond.)" = 2, "Amount (Cond.)" = 2))
```

The treatment group shows a notably higher response rate than the control group, and total donations are also higher. The differences across match ratios (1:1 vs. 2:1 vs. 3:1) are small and largely indistinguishable from noise — a key finding of the paper. Notably, the conditional donation amount, the average gift among donors only, is slightly lower in the treatment group ($43.87) than in the control ($45.54). This suggests the matching offer brings in more donors at the margin, but those marginal donors tend to give smaller amounts.

Below is a conditional vs. unconditional donation amounts by group chart
```{r}
#| label: fig-amounts
#| fig-cap: "Average donation amount by group. Unconditional includes non-donors (amount = 0); conditional is among donors only. The treatment group attracts more donors but at slightly lower average gift sizes."

amount_plot <- df |>
  filter(!is.na(ratio) | control == 1) |>
  mutate(group = case_when(
    control == 1 ~ "Control",
    ratio == 1   ~ "1:1 Match",
    ratio == 2   ~ "2:1 Match",
    ratio == 3   ~ "3:1 Match"
  )) |>
  filter(!is.na(group)) |>
  group_by(group) |>
  summarise(
    unconditional = mean(amount, na.rm = TRUE),
    conditional   = mean(amount[gave == 1], na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(group = factor(group, levels = c("Control", "1:1 Match", "2:1 Match", "3:1 Match"))) |>
  pivot_longer(c(unconditional, conditional), names_to = "type", values_to = "amount") |>
  mutate(type = recode(type,
    "unconditional" = "Unconditional (all recipients)",
    "conditional"   = "Conditional (donors only)"
  ))

ggplot(amount_plot, aes(x = group, y = amount, fill = group, alpha = type)) +
  geom_col(position = position_dodge(width = 0.6), width = 0.5, color = "white") +
  scale_alpha_manual(values = c("Unconditional (all recipients)" = 0.5,
                                "Conditional (donors only)" = 1.0),
                     name = NULL) +
  scale_fill_manual(values = c("Control"   = "#9E9E9E",
                               "1:1 Match" = "#1565C0",
                               "2:1 Match" = "#1976D2",
                               "3:1 Match" = "#42A5F5"),
                    guide = "none") +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(
    x = NULL,
    y = "Average donation ($)",
    title = "Matching Raises Participation, Not Gift Size",
    subtitle = "Conditional vs. unconditional donation amounts by group (Karlan & List 2007)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title    = element_text(face = "bold"),
    plot.subtitle = element_text(color = "grey50"),
    legend.position = "bottom"
  )
```

## Table 3: Effect of Matching on Likelihood of Donation (OLS)

The paper uses a probit model (reporting marginal effects via `dprobit`) to estimate how the matching grant affects the probability of donating. Here I replicate those results using **OLS with robust standard errors** (a linear probability model). The OLS coefficient on `treatment` is directly interpretable as a marginal effect on the probability of giving, and in large samples with a binary outcome it closely approximates probit marginal effects.

I replicate two specifications:

- **Column 1**: Bivariate — just the treatment indicator.
- **Column 2**: With controls for match ratio dummies, match threshold size dummies, and suggested ask-amount dummies.

```{r}
#| label: table3

# Robust SE helper
robust_se <- function(model) {
  coeftest(model, vcov = vcovHC(model, type = "HC1"))
}

m1 <- lm(gave ~ treatment, data = df)
m2 <- lm(gave ~ treatment + ratio2 + ratio3 + size25 + size50 + size100 + askd2 + askd3,
         data = df)

r1 <- robust_se(m1)
r2 <- robust_se(m2)

# Extract rows of interest
extract_coef <- function(r, vars) {
  map_dfr(vars, function(v) {
    if (v %in% rownames(r)) {
      tibble(
        Variable = v,
        Estimate = round(r[v, "Estimate"], 5),
        `Std. Error` = round(r[v, "Std. Error"], 5),
        `p-value` = round(r[v, "Pr(>|t|)"], 3)
      )
    } else {
      tibble(Variable = v, Estimate = NA, `Std. Error` = NA, `p-value` = NA)
    }
  })
}

vars_col1 <- c("(Intercept)", "treatment")
vars_col2 <- c("(Intercept)", "treatment", "ratio2", "ratio3",
               "size25", "size50", "size100", "askd2", "askd3")

col1 <- extract_coef(r1, vars_col1) |> rename_with(~ paste0(.x, " (1)"), -Variable)
col2 <- extract_coef(r2, vars_col2) |> rename_with(~ paste0(.x, " (2)"), -Variable)

t3 <- full_join(col1, col2, by = "Variable") |>
  mutate(Variable = recode(Variable,
    "(Intercept)" = "Intercept",
    "treatment"   = "Treatment",
    "ratio2"      = "Match ratio 2:1",
    "ratio3"      = "Match ratio 3:1",
    "size25"      = "Match threshold $25k",
    "size50"      = "Match threshold $50k",
    "size100"     = "Match threshold $100k",
    "askd2"       = "Ask amount: medium",
    "askd3"       = "Ask amount: high"
  ))

kable(
  t3,
  caption = "Table 3 (Panel A): OLS Estimates of Effect on Probability of Giving",
  align   = c("l", rep("r", 6))
) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) |>
  add_header_above(c(" " = 1, "Column (1): Bivariate" = 3, "Column (2): With Controls" = 3))
```

```{r}
#| echo: false
trt_coef1 <- round(coef(m1)["treatment"] * 100, 2)
trt_coef2 <- round(coef(m2)["treatment"] * 100, 2)
```

The coefficient on `treatment` in the bivariate model is approximately **`r trt_coef1` percentage points** — meaning recipients who received a matching grant offer were about `r trt_coef1` pp more likely to donate. Adding controls in Column 2 yields a smaller estimate of **`r trt_coef2` pp**, about half the bivariate estimate, though both are directionally consistent.


## Visualization: Response Rate by Match Ratio

One of the paper's surprising findings is that response rates do not increase with the generosity of the match. The plot below makes this pattern vivid.

```{r}
#| label: fig-ratios
#| fig-cap: "Response rate (share donating) by experimental group. Error bars are ±1 SE. The dashed line marks the control group rate. Larger match ratios do not meaningfully outperform a 1:1 match."

plot_data <- df |>
  filter(!is.na(ratio) | control == 1) |>
  mutate(
    group = case_when(
      control == 1 ~ "Control",
      ratio == 1   ~ "1:1 Match",
      ratio == 2   ~ "2:1 Match",
      ratio == 3   ~ "3:1 Match"
    )
  ) |>
  filter(!is.na(group)) |>
  group_by(group) |>
  summarise(
    rate = mean(gave, na.rm = TRUE),
    se   = sd(gave, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  ) |>
  mutate(group = factor(group, levels = c("Control", "1:1 Match", "2:1 Match", "3:1 Match")))

ctrl_rate <- plot_data$rate[plot_data$group == "Control"]

ggplot(plot_data, aes(x = group, y = rate, fill = group)) +
  geom_col(width = 0.55, color = "white", alpha = 0.85) +
  geom_errorbar(aes(ymin = rate - se, ymax = rate + se),
                width = 0.15, linewidth = 0.7, color = "grey30") +
  geom_hline(yintercept = ctrl_rate, linetype = "dashed",
             color = "grey40", linewidth = 0.6) +
  annotate("text", x = 0.6, y = ctrl_rate + 0.0003,
           label = "Control rate", hjust = 0, size = 3, color = "grey40") +
  scale_fill_manual(values = c("Control" = "#9E9E9E",
                               "1:1 Match" = "#1565C0",
                               "2:1 Match" = "#1976D2",
                               "3:1 Match" = "#42A5F5")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1),
                     expand = expansion(mult = c(0, 0.08))) +
  labs(
    x = NULL,
    y = "Share of recipients who donated",
    title = "Matching Grants Raise Giving — But Bigger Matches Don't Help More",
    subtitle = "Response rates by experimental group (Karlan & List 2007)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    plot.title    = element_text(face = "bold"),
    plot.subtitle = element_text(color = "grey50")
  )
```

All three match-ratio groups outperform the control, but the response rates for 2:1 and 3:1 are essentially indistinguishable from 1:1. This is the paper's central message on ratio: **announcing a match matters; the size of the match does not**. The implication for fundraisers is that a generous but expensive 3:1 match is no more effective at the margin than a cheaper 1:1 offer.


## Conclusion

This replication confirms the core findings of Karlan & List (2007):

1. **Matching grants work.** Simply offering to match donations raises the response rate by roughly 0.41 percentage points (about 23% above the control baseline) and increases total revenues.
2. **Match ratio is irrelevant at the margin.** There is no detectable gain from increasing the ratio from 1:1 to 2:1 or 3:1 — the "announcement effect" saturates quickly.
3. **Randomization was successful.** Table 1 shows no statistically significant pre-treatment differences between groups, validating the causal interpretation of all results.

Together, these results suggest that charitable giving responds to price changes, but that donors are not as responsive to the *magnitude* of the price reduction as a simple price-elasticity model would predict — pointing toward behavioral mechanisms such as social proof or signals of organizational quality.
